Adaptive evolution with aneuploidy and mutation

### 2022-01-05 ms/ms.tex
added heading for analysis results
added convergence figures; added convergence.ipynb

\documentclass[12pt]{extarticle}
\usepackage{geometry}
\geometry{
a4paper,
total={170mm,257mm},
left=20mm,
top=20mm,
headheight=12pt
}

\usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx} % Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
% TeX will automatically convert eps --> pdf in pdflatex

@@

\paragraph{Likelihood function.} 
% TODO please read!
Because our model, just like the Wright-Fisher model, is non-linear and stochastic, computing the distribution of fixation time $T(g)$ of genotype $g$ for use in the likelihood function is intractable (it is even hard to use a diffusion-equation approximation due to the model having multiple genotypes, rather than just two).
We overcome this problem by approximating the likelihood using simulations. We simulate 1,000 experiments per parameter vector $\theta = (\mu, \tau, \delta, s, b, c)$, resulting in a set of simulated observations $\tilde{\vec X} = \{\tilde{X}_i\}_{i=1}^{1000}$. We then compute the approximate likelihood,
\begin{equation}\begin{aligned}
\label{eq:heatstress-likelihood}
\likelihood(\theta) = &\ P^4(200 \le T(2n+1) \le 450) \cdot 
	\Big[1 - \\
	&	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\} \mid 200 \le T(2n+1) \le 450\big)- \\
	&	P_{\tilde{\vec X}}^4\big(!\{1700 < T(2n^*) < 2350\} \mid 200 \le T(2n+1) \le 450\big)+ \\
	&	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\} \land !\{1700 < T(2n^*) < 2350\} \mid 200 \le T(2n+1) \le 450\big) 
	\Big]\;,
\end{aligned}\end{equation}
where $!\{\ldots\}$ is the "logical not" operator, $P^4(\ldots)$ is the 4th power of $P(\ldots)$, and all probabilities $P_{\tilde{\vec X}}(\ldots)$ are approximated from the results of the simulations $\tilde{\vec X}$. For example, $P_{\tilde{\vec X}}\big(!\{T(2n^*)<1700\} \mid 200 \le T(2n+1) \le 450\big)$ is approximated by taking simulations in which \anwt\; fixed before generation 450 but not before generation 200, and computing the fraction of such simulations in which \eumt\; did not fix by generation 1,700, and hence aneuploidy did not extinct before generation 1,700.

@@

Acceptance is determined according to the approximate likelihood, which has a maximum value of 0.875. Thus, we terminated the inference when $\epsilon \le 0.13$ after four iterations, with $n_4=1,074$ accepted parameter vectors and effective sample size (ESS) 700 (\autoref{fig:ess-single}). % TODO for multi-locus model we need to say something else
Running the inference algorithm with different initialization seeds and different number of simulations with produced similar outcomes (\autoref{fig:seeds}). %TODO for multi-locus model, write about them here also.

@@

% Fig ABC convergence
\begin{figure}[h]
    \centering
	\includegraphics[width=1\textwidth]{../figures/ess.pdf}
    \caption{
    \textbf{Single-locus model inference convergence.} Measures of convergence for the ABC-SMC algorithm with the single-locus model. After iteration number 4, the acceptance threshold was $\epsilon=0.13$, the acceptance rate was XXX, the number of accepted parameter sets ("particles") was 1,074, and the effective sample size (ESS) 700. % TODO add missing acceptance rate 
    } 
    \label{fig:ess-single}
    % TODO y label in first panel "Acceptance threshold, $\epsilon$"
    % add panel labels A B C D
    % x label - just once
	% should we do 1-3 more iterations? it is not clear that it converged.
\end{figure}
% TODO show posteriors of different iterations (with panel labels!!)

>>>>

\documentclass[12pt]{extarticle}
\usepackage{geometry}
\geometry{
a4paper,
total={170mm,257mm},
left=20mm,
top=20mm,
headheight=12pt
}
\usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx} % Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
% TeX will automatically convert eps --> pdf in pdflatex

@@

\paragraph{Likelihood function.} 
Because our model, just like the Wright-Fisher model, is non-linear and stochastic, computing the distribution of fixation time $T(g)$ of genotype $g$ for use in the likelihood function is intractable (it is even hard to use a diffusion-equation approximation due to the model having multiple genotypes, rather than just two).
We overcome this problem by approximating the likelihood using simulations. We simulate 1,000 experiments per parameter vector $\theta = (\mu, \tau, \delta, s, b, c)$, resulting in a set of simulated observations $\tilde{\vec X} = \{\tilde{X}_i\}_{i=1}^{1000}$. We then compute the approximate likelihood,
\begin{equation}\begin{aligned}
\label{eq:heatstress-likelihood}
\likelihood(\theta) = &\ P^4(200 \le T(2n+1) \le 450) \cdot 
	\Big[1 - \\
	&	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\} \mid 200 \le T(2n+1) \le 450\big)- \\
	&	P_{\tilde{\vec X}}^4\big(!\{1700 < T(2n^*) < 2350\} \mid 200 \le T(2n+1) \le 450\big)+ \\
	&	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\} \land !\{1700 < T(2n^*) < 2350\} \mid 200 \le T(2n+1) \le 450\big) 
	\Big]\;,
\end{aligned}\end{equation}
where $!\{\ldots\}$ is the "logical not" operator, $P^4(\ldots)$ is the 4th power of $P(\ldots)$, and all probabilities $P_{\tilde{\vec X}}(\ldots)$ are approximated from the results of the simulations $\tilde{\vec X}$. For example, $P_{\tilde{\vec X}}\big(!\{T(2n^*)<1700\} \mid 200 \le T(2n+1) \le 450\big)$ is approximated by taking simulations in which \anwt\; fixed before generation 450 but not before generation 200, and computing the fraction of such simulations in which \eumt\; did not fix by generation 1,700, and hence aneuploidy did not extinct before generation 1,700.

@@

Acceptance is determined according to the approximate likelihood, which has a maximum value of 0.875. Thus, we terminated the inference when $\epsilon \le 0.13$ after six iterations, with $n_6=982$ accepted parameter vectors and effective sample size (ESS) 651 (\autoref{fig:convergence}). % TODO for multi-locus model we need to say something else
Running the inference algorithm with different initialization seeds and different number of simulations with produced similar outcomes (\autoref{fig:seeds}). %TODO for multi-locus model, write about them here also.

@@

% Fig ABC iterations
\begin{figure}[h]

  \begin{subfigure}{1\textwidth}
    \centering
      \includegraphics[width=0.45\textwidth]{../figures/convergence-p1_mr.pdf}      
      \includegraphics[width=0.45\textwidth]{../figures/convergence-p2_tr.pdf}      
      \includegraphics[width=0.45\textwidth]{../figures/convergence-p3_w1.pdf}      
      \includegraphics[width=0.45\textwidth]{../figures/convergence-p4_w2.pdf}      
      \includegraphics[width=0.45\textwidth]{../figures/convergence-p5_w3.pdf}      
     \label{fig:convergence-A}
    \caption{\textbf{A. Posterior densities per iteration}}
    \end{subfigure}
    \\
    \vspace*{4mm}

  \begin{subfigure}{1\textwidth}
  	\includegraphics[width=1\textwidth]{../figures/ess.pdf}
     \label{fig:convergence-B}
     \vspace*{-8mm}
    \caption{\textbf{B. ABC-SMC population properties per iteration}}
  \end{subfigure}

    \caption{
    \textbf{Single-locus model inference convergence.} 
    ABC-SMC algorithm was executed to inference single-locus model parameters. \textbf{(A)} For each parameter of the model, there's shown KDE of the posterior density of the inference per iteration.
We can see that the posterior didn't change significantly already from the first iteration. \textbf{(B)} Here are shown measures of convergence for the ABC-SMC algorithm. After iteration number 6, the acceptance threshold was $\epsilon=0.13$, the acceptance rate was $0.018$, the number of particles was 982, and the effective sample size (ESS) 651.
}
    \label{fig:convergence}
\end{figure}
